---
title: "r coding club mixed models intro"
author: "Stefan Van Dongen"
date: "10 september 2018"
output:
  slidy_presentation: default
#  ioslides_presentation: default
#  beamer_presentation: default
# fig_width: 6 
# fig_height: 4 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Mixed/hierarchical models

1. Why?
    + Because you often have to!
    + Whenever there is some (nuisance)-structure in your data
2. What?
    + Models with random effects!
3. How? 
    + Easy in R (or many other packages)
4. What is then so difficult about it? 
    + Set up your model
    + (Interpret the output) 

## Why?
* Let us talk about the radon data
* Radon radiation in houses (878) in different counties of Minnesota (85 counties)
* Some houses have a basement, others do not (with basement one expects higher radon issue)
* at county level, ground-uranium levels are known (may allow predictions)


## The radon dataset

```{r radondata, echo = F,comment=NA}
rad<-read.table('radon.txt',header=T)
attach(rad)
head(rad)
table(basement)
```

## Does the amount of Uranium predict radon radiation in houses and are there differences in buildings with or without basement

```{r plot radon vs uranium 1,comment=NA}
plot(uranium,radon,pch=as.integer(basement),col=as.integer(basement))
```

## A simple regression

```{r simple regression,comment=NA}
summary(lm(radon~uranium))
```

## Data not independent at county level

```{r plot with averages,comment=NA}
library(Hmisc,warn.conflicts=F, quietly=T)
means_r<-tapply(radon,county,mean)
sd<-tapply(radon,county,sd)
n<-tapply(radon,county,length)
se_r<-sd/sqrt(n)
means_u<-tapply(uranium,county,mean)
errbar(means_u,means_r,yplus=(means_r+se_r),ymin=(means_r-se_r))
```


## A simple regression on the averages?

```{r simple regression2,comment=NA}
summary(lm(means_r~means_u))
```
## Add county as fixed effect?

```{r anova1,comment=NA}
summary(lm(radon~uranium+county))
```

## A mixed model!

```{r mixed model1,comment=NA, echo=TRUE, message=F, warning=F}
library(lmerTest)
lm1<-lmer(radon~uranium+(1|county))
summary(lm1)
```

## What is the random effects part?

```{r mixed model2,comment=NA}
head(data.frame(ranef(lm1)),n=10)
```


## What is the random effects part?

```{r mixed model3,comment=NA}
errbar(means_u,means_r,yplus=(means_r+se_r),ymin=(means_r-se_r))
abline(0.55,0.775)
points(means_u[7],means_r[7],col='blue',cex=1,pch=19)
```

## making predictions

```{r predictions,comment=NA}
errbar(means_u,means_r,yplus=(means_r+se_r),ymin=(means_r-se_r))
abline(0.55,0.775)
points(means_u[7],means_r[7],col='blue',cex=1,pch=19)
points(means_u[7],predict(lm1,newdata=data.frame(uranium = means_u[7]),re.form=NA),cex=1,pch=19,col='red')
points(means_u[7],predict(lm1,newdata=data.frame(uranium = means_u[7], county='BLUEEARTH'),re.form=~(1|county)),cex=1,pch=19,col='green')
```

## Add basement effect

```{r mixed with basement1,comment=NA, echo=TRUE, message=F, warning=F}
lm2<-lmer(radon~uranium+basement+(1|county))
summary(lm2)
```

## Random slopes needed

```{r mixed with basement2,comment=NA, echo=TRUE, message=F, warning=F}
lm3<-lmer(radon~uranium+basement+(1+basement|county))
summary(lm3)
```

## Random slopes needed 2

```{r mixed with basement3,comment=NA, echo=TRUE, message=F, warning=F}
base<-(basement == 'Y')-0.5
lm3<-lmer(radon~uranium+base+(1+base|county))
summary(lm3)
```

## Random slopes interpretation

```{r mixed with basement4}
int_slopes<-ranef(lm3)$county
coeff<-coef(lm3)$county
means_rn<-tapply(radon[basement == 'N'],county[basement == 'N'],mean)
means_ry<-tapply(radon[basement == 'Y'],county[basement == 'Y'],mean)
means_r2<-c(means_rn[1:7],means_ry[1:7])
basem<-c(rep(-0.5,7),rep(0.5,7))
plot(basem,means_r2,pch=19,ylab='Radon levels',xlab='basement no or yes',col=c(1:7,1:7),xlim=c(-0.75,0.75))
for(i in 1:5){
  lines(c(basem[i],basem[i+7]),c(means_r2[i],means_r2[i+7]),col=i,lwd=4)}
for(i in 1:7){
  abline((coeff[i,1]+coeff[i,2]*means_u[i]),coeff[i,3],col=i,lwd=2,lty=2)
}


```

## Let us all become Bayesians ;-)

1. Why Bayesian?
    + it makes more sence
    + it is very flexible!!
    + include prior information
2. Why not Bayesian?
    + Faster
    + non-parametrics often easy
    + No need to chose a prior
    + More diagnostics


## Radon data analysis in a Bayesian model

```{r bayesian1,comment=NA, echo=TRUE, message=F, warning=F,tidy=TRUE}
library(MCMCglmm)
x<-data.frame(cbind(radon,uranium,base,county))
lm4<-MCMCglmm(radon~uranium+base,random=~county+base:county,data=x,verbose=F)
plot(lm4$VCV)
```

## The Bayesian analysis with a longer MCMC

```{r bayesian2,comment=NA, echo=TRUE, message=F, warning=F,tidy=T}
lm5<-MCMCglmm(radon~uranium+base,random=~county+base:county,data=x,nitt=550000,thin=50,burnin=50000,verbose=F)
plot(lm5$VCV)
```

## Interpretation of the output

```{r bayesian3,comment=NA, echo=TRUE, message=F, warning=F}
summary(lm5)
```

## For the 'real' Bayesians: OpenBugs

```{r bayesian figure, message=F, warning=F}
library(png)
library(gridGraphics)
img <- readPNG("C:/Users/Support/Documents/r coding club/images/winbugs1.png")
grid.raster(img)
```


